Internal models of type theory and internal higher categories for the Univalent Foundations

### 2014-03-04 article-models-tt.tex
outlined the structure of the type constructors section
starting to work on models

\section{Structural type theory}\label{tt}
In this section we give a description of type theory before basic
constructors and with explicit extension, weakening and substitution. We will
formulate the type theory in such a way that contexts aren't lists of variable
declarations. The main reason is that we don't see variable declarations in the
models either. This way we also set out to a more algebraic approach of type
theory and higher category theory. Thirdly, we will not have to be burdened with
comments about variables being bounded or not, or fresh or not occuring at all.

@@

@@

\begin{proof}
Let $\jhom{\Gamma}{A}{B}{f}$, $\jhom{\Gamma}{B}{C}{g}$, $\jfam{{\Gamma}{C}}{R}$
and $\jterm{{\Gamma}{B}}{\subst{g}{\ctxwk{B}{R}}}{w}$. Then we have the
judgmental equalities
\begin{align*}
\jcomp{f}{\ctxext{g}{w}}
& \jdeq \subst{f}{\ctxwk{A}{\ctxext{g}{w}}}\\
& \jdeq \subst{f}{\ctxext{\ctxwk{A}{g}}{\ctxwk{A}{w}}}\\
& \jdeq \ctxext{\subst{f}{\ctxwk{A}{g}}}{\subst{f}{\ctxwk{A}{w}}}\\
& \jdeq \ctxext{\jcomp{f}{g}}{\jcomp{f}{w}}.
\end{align*}
\end{proof}

@@

>>>>

\section{Type theory before type constructors}\label{tt}
In this section we give a description of type theory before basic
constructors and with explicit extension, weakening and substitution. We will
formulate the type theory in such a way that contexts aren't lists of variable
declarations. The main reason is that we don't see variable declarations in the
models either. This way we also set out to a more algebraic approach of type
theory and higher category theory. Thirdly, we will not have to be burdened with
comments about variables being bounded or not, or fresh or not occuring at all.

@@

There is also a notion of morphism \emph{over} a morphism. We will develop this
notion because it will be needed in the theory of models later on.

\begin{defn}
Let $\jhom{\Gamma}{A}{B}{f}$ be a morphism from $A$ to $B$ in context $\Gamma$
and consider $\jfam{{\Gamma}{A}}{P}$ and $\jfam{{\Gamma}{B}}{Q}$. A morphism
from $P$ to $Q$ over $f$ in context $\Gamma$ is a term
\begin{equation*}
\jfhom{\Gamma}{A}{B}{f}{P}{Q}{F}.
\end{equation*}
\end{defn}

We can also define composition for this version of morphism and formulate
compatibility rules for them.

@@

\begin{proof}
Let $\jhom{\Gamma}{A}{B}{f}$, $\jhom{\Gamma}{B}{C}{g}$, $\jfam{{\Gamma}{C}}{R}$
and $\jterm{{\Gamma}{B}}{\subst{g}{\ctxwk{B}{R}}}{w}$. Then we have the
judgmental equalities
\begin{align*}
\jcomp{f}{\ctxext{g}{w}}
& \jdeq \subst{f}{\ctxwk{A}{\ctxext{g}{w}}}\\
& \jdeq \subst{f}{\ctxext{\ctxwk{A}{g}}{\ctxwk{A}{w}}}\\
& \jdeq \ctxext{\subst{f}{\ctxwk{A}{g}}}{\subst{f}{\ctxwk{A}{w}}}\\
& \jdeq \ctxext{\jcomp{f}{g}}{\jcomp{f}{w}}.
\end{align*}
\end{proof}
%\end{document}

\section{Type constructors in structural type theory}
In this section we will introduce the usual type constructors to structural
type theory. We do so with the point of view that each type constructor is a
(class of) operations on type theory that should be compatible with extension,
weakening, substitution and identity morphisms. Likewise, each induction
principle of an inductively defined type constructor is going to be such an
operator, and it is therefore also required to be compatible with extension,
weakening, substitution and identity functions.

We will first state all the rules of the individual type constructors, where
only the $\wtypesym$-type constructor depends on the presence of dependent
function types (or universes too?), and hence comes with the notion that $\wtypesym$
is compatible with dependent product types and vice versa. All the other
compatibility properties are stated in \autoref{compatibility-of-type-constructors}.

\subsubsection{Issues to keep in mind}
\begin{enumerate}
\item The rule asserting that unit types are compatible with context extension makes
no sense from the point of view that contexts are lists of types and types
are contexts of length one; in other words, that types are indecomposable (i.e.~non-extended)
families. Remidies:
\begin{enumerate}
\item This point of view is wrong and should be abandoned.
\item We have, as Vladimir proposed, two kinds of judgmental equalities. One
      could be used for the very strict equalities, the other could be used
      to state the compatibility rules with. In other words, the other equality
      is a compatibility relation. If we do that, we shouldn't require that
      if a family is compatible with a type, then the family is a type. The
      compatibility relation could relate things by uniqueness up to unique
      isomorphism.
\item Don't state such compatibility rules for the type constructors.
\item Don't even give the unit type a dependent action.
\end{enumerate}
\end{enumerate}

@@

\begin{align*}
& \inference
  {\jfam{\Gamma}{A}}
  {\jtype{{\Gamma}{A}}{\unitc{A}}}
& & \inference
    {\jfameq{\Gamma}{A}{A'}}
    {\jtypeeq{{\Gamma}{A}}
      {\unitc{A}}
      {\unitc{A'}}}\\
& \inference
  {\jfam{\Gamma}{A}}
  {\jterm{{\Gamma}{A}}{\unitc{A}}{\unitct{A}}}
& & \inference
    {\jfameq{\Gamma}{A}{A'}}
    {\jtermeq{{\Gamma}{A}}{\unitc{A}}{\unitct{A}}{\unitct{A'}}}\\
& \inference
  {\jfam{{\Gamma}{A}}{P}}
  {\jtype{{{{\Gamma}{A}}{P}}{\ctxwk{P}{\unitc{A}}}}{\unitf{A}{P}}}
& & \inference
    {\jfameq{{\Gamma}{A}}{P}{P'}}
    {\jtypeeq{{{{\Gamma}{A}}{P}}{\ctxwk{P}{\unitc{A}}}}
      {\unitf{A}{P}}
      {\unitf{A}{P'}}}\\
& \inference
  {\jfam{{\Gamma}{A}}{P}}
  {\jterm{{{{\Gamma}{A}}{P}}{\ctxwk{P}{\unitc{A}}}}{\unitf{A}{P}}{\unitft{A}{P}}}
& & \inference
    {\jfameq{{\Gamma}{A}}{P}{P'}}
    {\jtermeq{{{{\Gamma}{A}}{P}}{\ctxwk{P}{\unitc{A}}}}{\unitf{A}{P}}
      {\unitft{A}{P}}
      {\unitft{A}{P'}}}
\end{align*}

We impose the following compatibility rules for the unit type:

\begin{align*}
& \inference
  {\jfam{{\Gamma}{A}}{P}}
  {\jtypeeq{{{\Gamma}{A}}{P}}{\ctxext{\ctxwk{P}{\unitc{A}}}{\unitf{A}{P}}}{\unitc{{\Gamma}{A}}}}\\
& \inference
  {\jfam{\Gamma}{A}
   \jfam{\Gamma}{B}}
  {\jtypeeq{{{\Gamma}{A}}{\ctxwk{A}{B}}}{}{}}  
\end{align*}

\subsection{The empty type}

\subsection{The natural numbers}

\subsection{Dependent pair types}

\subsection{Identity types}

\subsection{Dependent function types}

\subsection{$\wtypesym$-types}

\subsection{Universes}

\subsection{The compatibility of the type constructors with each other}
\label{compatibility-of-type-constructors}

\section{Models of type theory}
In this section we pursue the idea of what a general model of type theory is by
axiomatizing what you can do with them. We have the following ideas:
\begin{enumerate}
\item There are dependent models and sections thereof. Particular instances
of sections: extension, weakening and substitution (and something for identity
function?). And like the original
extension, weakening and substitution, they're going to be compatible with each
other. Therefore we state our theory of models as an extension of type theory
without constructors.
\item Type theory without constructors is a model of itself, the canonical model $\mctx$.
\item for any (family of) model(s) $A$, there is the family model $\mfam{A}$ which
is a family of models over $A$.
\item The terms of $\mctx$ should be precisely contexts. Terms of $\subst{\Gamma}{\mfam{\mctx}}$
should be families over $\Gamma$.
\item if a model and a family of models over it are given, there is an extended model.
If we extend $A$ by $\mfam{A}$, we get the Sierpinski model of $A$.
\item Likewise, models can be weakened and substituted and there are identity
functions.

\end{enumerate}
So the theory of models is going to be an extension of this theory with this
data.

\subsection{The basic ingredients of the theory of models}
We first introduce the basic ingredients of our abstract theory of models.
\begin{align*}
& \inference
  {}
  {\jctx{\mctx}}
  \tag{the canonical model}\\
& \inference
  {\jfam{\Gamma}{A}}
  {\jfam{{\Gamma}{A}}{\mfam{A}}}
  \tag{families}\\
& \inference
  {\jterm{\Gamma}{A}{x}}
  {\jterm{{\Gamma}{A}}{\mfam{A}}{\mfam{x}}}
  \tag{term action of $\mfam{}$}\\
& \inference
  {\jctx{\Gamma}}
  {\jfam{{\Gamma}{\mfam{\Gamma}}}{\mtm{\Gamma}}}
  \tag{terms}
\intertext{The empty context of a model $A$ is simply a term of $A$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jterm{\Gamma}{A}{\tfemp{A}}}
  \tag{empty context}
\intertext{Extension is going to be a morphism from $\ctxext{A}{\mfam{A}}$ to
$A$ in context $\Gamma$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jhom{\Gamma}{{A}{\mfam{A}}}{A}{\tfext{A}}}
  \tag{extension}
\intertext{The action of weakening on families
is a morphism from $\ctxwk{\mfam{A}}{\mfam{A}}$ to
$\mfam{\mfam{A}}$ in context $\ctxext{{\Gamma}{A}}{\mfam{A}}$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jhom{{{\Gamma}{A}}{\mfam{A}}}{\ctxwk{\mfam{A}}{\mfam{A}}}
    {\mfam{\mfam{A}}}
    {\tfwk{A}^0}}
  \tag{weakening}
\intertext{The action of weakening on terms
is a morphism from $\ctxwk{\mfam{A}}{\mtm{A}}$ to $\mtm{\mfam{\mfam{A}}}$ over
$\tfwk{A}^0$ in context $\ctxext{{\Gamma}{A}}{\mfam{A}}$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jfhom
    {{{\Gamma}{A}}{\mfam{A}}}
    {\ctxwk{\mfam{A}}{\mfam{A}}}
    {\mfam{\mfam{A}}}
    {\tfwk{A}^0}
    {\ctxwk{\mfam{A}}{\mtm{A}}}
    {\mtm{\mfam{\mfam{A}}}}
    {\tfwk{A}^1}}
  \tag{weakening}
\intertext{The action of substitution by a term on families
is going to be a morphism from $\ctxwk{\mtm{A}}{\mfam{\mfam{A}}}$ to 
$\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}$
in context $\ctxext{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jhom
  {{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}}
  {\ctxwk{\mtm{A}}{\mfam{\mfam{A}}}}
  {\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}}
  {\tfsubst{A}^0}}
\intertext{Likewise, the action of substitution by a term on terms of those
families is going to be a morphism from $\ctxwk{\mtm{A}}{\mtm{\mfam{\mfam{A}}}}$
to $\ctxwk{\mtm{A}}{{\mfam{A}}{\mtm{A}}}$ over $\tfsubst{A}^0$ in context
$\ctxext{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jfhom
  {{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}}
  {\ctxwk{\mtm{A}}{\mfam{\mfam{A}}}}
  {\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}}
  {\tfsubst{A}^0}
  {\ctxwk{\mtm{A}}{\mtm{\mfam{\mfam{A}}}}}
  {\ctxwk{\mtm{A}}{{\mfam{A}}{\mtm{A}}}}
  {\tfsubst{A}^1}}
\intertext{The identity functions are coded by}
& \inference
  {\jfam{\Gamma}{A}}
  {\jterm
   {{{\Gamma}{A}}{\mfam{A}}}
   {\subst{{\idfunc[\mfam{A}]}{\tfwk{A}^0}}{\mtm{\mfam{\mfam{A}}}}}
   {\tfid{A}}}
\end{align*}
We will call the terms that we introduced here the model constructors.

\subsection{The compatibility rules}
We need to do several things in this section:
\begin{enumerate}
\item postulate that ordinary extension, weakening, substitution and identity
functions are compatible with the model constructors.
\item postulate that each of the model constructors is compatible with ordinary
extension, weakening, substitution and identity functions.
\item postulate that the model constructors are compatible with each other,
so that they come to model ordinary extension, weakening, substitution and
identity functions respectively.
\end{enumerate}
Note that the rules for compatibility with extension are going to explain which
model $\ctxext{A}{P}$ is by telling what the families, the terms, extension,
weakening, substitution and identity functions are. Likewise the rules for
compatibility with weakening and substitution do this for their respective
cases.

\subsubsection{Families of extensions}
\begin{equation*}
\inference
  {\jfam{{\Gamma}{A}}{P}}
  {\jfameq{{{\Gamma}{A}}{P}}
    {\mfam{{A}{P}}}
    {{\ctxwk{P}{\mfam{A}}}{\ctxwk{{P}{\mfam{A}}}{\mfam{P}}}}}
\end{equation*}

\subsubsection{Terms of families of extensions}
\begin{equation*}
\inference
  {\jterm{\Gamma}{A}{x}
   \jterm{\Gamma}{\subst{x}{P}}{u}}
  {\jtermeq
    {{{\Gamma}{A}}{P}}
    {\mfam{{A}{P}}}
    {\mfam{{x}{u}}}
    {{\ctxwk{P}{\mfam{x}}}{\mfam{u}}}}
\end{equation*}

\subsubsection{Terms of extensions}
\begin{equation*}
\inference
  {\jfam{{\Gamma}{A}}{P}}
  {\jfameq
    {{{{\Gamma}{A}}{P}}{\mfam{{A}{P}}}}
    {\mtm{{A}{P}}}
    {{\ctxwk{{P}{\mfam{A}}}{{P}{\mtm{A}}}}{\ctxwk{{P}{\mfam{A}}}{\mtm{P}}}}}
\end{equation*}
